
# Exercises 3 {-}

## E3.1 `Bmi2` {-}
Add a return value `advice_text` to the `Bmi2` function we drafted in the lecture. The value of `advice_text` should be chosen from one of the four character strings below according to `bim_value`:

 - "below 18.5 – you're in the underweight range"
 - "between 18.5 and 24.9 – you're in the healthy weight range"
 - "between 25 and 29.9 – you're in the overweight range"
 - "between 30 and 39.9 – you're in the obese range"

```{r eval = FALSE}
Bmi2 <- function(weight, height) {
  # your code
}
bb <- Bmi2(weight = 90, height = 1.75)
bb[["value"]]
bb[["reduce_weight"]]
bb[["advice_text"]]
```


## E3.2 `||`, `|`, `&&`, `&` {-}

Execute the following code snippet and answer the two questions in the comment:
```{r eval = FALSE}
rm(u)
u # expect to see an error
# Error: object 'u' not found

# 1. Why do the next two expressions evaluate to different results and why?
TRUE || u
TRUE | u

# 2. Why do the next two expressions evaluate to different results and why?
FALSE && u
FALSE & u
```


## E3.3 `table` {-}

Given `t <- table(c("apple", "banana", "coconut", "apple", "apple", "banana"))`
   1. What is the base type of `t`?
   2. What attributes are associated with `t`?
   3. What do you know about each of `t`'s attributes?
   4. Which attribute(s) does `t` share with a matrix?
   

## E3.4 `apply` {-}

The `runif` function generates a vector of [uniformly distributed](https://en.wikipedia.org/wiki/Continuous_uniform_distribution) random numbers within a given interval. For example, `runif(10, min = 2, max = 3)` returns 10 random numbers between `2` and `3`.

  1. With the help of this function, create a 1000 x 6 matrix of uniformly distributed values between `5` and `6`. 
  2. Calculate the 6 column-wise averages of the 1000 x 6 matrix without using a for loop.




