# Exercises {-}

## Exercises 1 {#sec-exercises-1}

### E1.1 `IsMonth` {-}


Write a function `IsMonth` which takes a single numerical value as input and returns TRUE iff that value may represent a month of the year.

```{r eval = FALSE}
IsMonth <- function (x) {
    # your code goes here
}
# Check if your implementation is correct with the followings
IsMonth(12.9) # FALSE
IsMonth(12)   # TRUE
IsMonth(1)    # TRUE
IsMonth(-10)  # FALSE
```


### E1.2 `IsLeapYear` {-}

Write a function `IsLeapYear` which takes a single numerical value as input and returns TRUE iff that value may represent a leap year. A leap year occurs on any year evenly divisible by 4, but not on a century unless it is divisible by 400.

```{r eval = FALSE}
IsLeapYear <- function (x) {
    # your code goes here
}
# Check if your implementation is correct with the followings
IsLeapYear(2000) # TRUE
IsLeapYear(1900) # FALSE
```


### E1.3 `Bmi` {-}
Write a function `Bmi` which calculates the Body Mass Index (BMI) given weight (in kilograms) and height (in meters) as inputs.

$$bmi =\frac{weight}{height^2}$$

```{r eval = FALSE}
Bmi <- function(weight, height) {
  # your code goes here
}
# Check if your implementation is correct with the followings
Bmi(weight = 90, height = 1.75) # 29.38776
```

### E1.4 `FeetToMeters` {-}
Write a function `FeetToMeters` which converts a numeric value in feet to meters. Given
$$1 ~ feet = 0.3048 ~ meters$$

```{r eval = FALSE}
FeetToMeters <- function(ft) {
  # your code goes here
}
# if your code is correct, this expression should evaluate to TRUE
all.equal(FeetToMeters(10), 3.048) 
```

### E1.5 `SqFeetToSqMeters` {-}
Use your implementation of `FeetToMeters` to convert a numeric value from square feet to square meters.

```{r eval = FALSE}
SqFeetToSqMeters <- function(sq_ft) {
  # your code goes here
}
# if your code is correct, this expression should evaluate to TRUE
all.equal(SqFeetToSqMeters(100), 9.290304)
```




