# Exercises I

## Exercises 1: Writing functions {#sec-exercises-1}

### E1.1 `IsMonth` {-}


Write a function `IsMonth` which takes a single numerical value as input and returns TRUE iff that value may represent a month of the year.

```{r eval = FALSE}
IsMonth <- function (x) {
    # your code goes here
}
# Check if your implementation is correct with the followings
IsMonth(12.9) # FALSE
IsMonth(12)   # TRUE
IsMonth(1)    # TRUE
IsMonth(-10)  # FALSE
```


### E1.2 `IsLeapYear` {-}

Write a function `IsLeapYear` which takes a single numerical value as input and returns TRUE iff that value may represent a leap year. A leap year occurs on any year evenly divisible by 4, but not on a century unless it is divisible by 400.

```{r eval = FALSE}
IsLeapYear <- function (x) {
    # your code goes here
}
# Check if your implementation is correct with the followings
IsLeapYear(2000) # TRUE
IsLeapYear(1900) # FALSE
```


### E1.3 `Bmi` {-}
Write a function `Bmi` which calculates the Body Mass Index (BMI) given weight (in kilograms) and height (in meters) as inputs.

$$bmi =\frac{weight}{height^2}$$

```{r eval = FALSE}
Bmi <- function(weight, height) {
  # your code goes here
}
# Check if your implementation is correct with the followings
Bmi(weight = 90, height = 1.75) # 29.38776
```

### E1.4 `FeetToMeters` {-}
Write a function `FeetToMeters` which converts a numeric value in feet to meters. Given
$$1 ~ feet = 0.3048 ~ meters$$

```{r eval = FALSE}
FeetToMeters <- function(ft) {
  # your code goes here
}
# if your code is correct, this expression should evaluate to TRUE
all.equal(FeetToMeters(10), 3.048) 
```

### E1.5 `SqFeetToSqMeters` {-}
Use your implementation of `FeetToMeters` to convert a numeric value from square feet to square meters.

```{r eval = FALSE}
SqFeetToSqMeters <- function(sq_ft) {
  # your code goes here
}
# if your code is correct, this expression should evaluate to TRUE
all.equal(SqFeetToSqMeters(100), 9.290304)
```

## Exercises 2: Control flow {#sec-exercises-2}

### E2.1 `SumMultiples` {-}
[Problem 1, Project Euler](https://projecteuler.net/problem=1): If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write a function `SumMultiples` to find the sum of all the multiples of 3 or 5 below 1000.

```{r eval = FALSE}
SumMultiples <- function(factor_1 = 3, factor_2 = 5, threshold = 1000) {
  # your code goes here
}
# test your code with the following calls
SumMultiples(threshold = 10) # 23
SumMultiples() # 233168
```


### E2.2 `SumEvenFib` {-}
[Problem 2, Project Euler](https://projecteuler.net/problem=2): Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be:

$$1, 2, 3, 5, 8, 13, 21, 34, 55, 89, ...$$

By considering the terms in the Fibonacci sequence whose values do not exceed four million, find the sum of the even-valued terms.

```{r eval = FALSE}
SumEvenFib <- function(max_fib) {
  # your code goes here
}
# test your code with the following calls
SumEvenFib(4e6) # 4613732
```


### E2.3 Hadley, 6.6.1.1 {-}
[Hadley, 6.6.1.1](https://adv-r.hadley.nz/functions.html#exercises-17): Explain the following results:

```{r eval = FALSE}
sum(1, 2, 3)  # 6
mean(1, 2, 3) # 1

sum(1, 2, 3, na.omit = TRUE)  # 7
mean(1, 2, 3, na.omit = TRUE) # 1
```

### E2.4 Hadley, 6.5.4.3 {-}
[Hadley, 6.5.4.3](https://adv-r.hadley.nz/functions.html#exercises-16): 
Without running the following code, answer the questions in the comment:

```{r eval = FALSE}
y <- 10
f1 <- function(x = {y <- 1; 2}, y = 0) {
  return(x + y)
}
x <- f1() # 1. what is the value of x here?
y         # 2. what is the value of y here?
```


### E2.5 `IsPalindrome` {-}
A palindrome is a sequence of characters which reads the same backward as forward. A number is a palindrome if its value is the same if it were written backwards. For example, 121, 53935, 11 are all palindrome numbers.

Write a function `IsPalindrome` which takes an integer `x` as input and returns `TRUE` iff it is a palindrome.

```{r eval = FALSE}
IsPalindrome <- function(x) {
  # your code
}
stopifnot(
  IsPalindrome(0)    &&
  IsPalindrome(1)    &&
  IsPalindrome(11)   &&
  IsPalindrome(121)  &&
  IsPalindrome(5412145)  &&
  !IsPalindrome(-121) &&
  !IsPalindrome(1231) &&
  !IsPalindrome(123)
)
print("Passed all tests")
```


### E2.6 `MaxPalindrome` {-}
[Problem 4, Project Euler](https://projecteuler.net/problem=4) The largest palindrome made from the product of two 2-digit numbers is 9009 = 91 × 99. Find the largest palindrome made from the product of two 3-digit numbers.

```{r eval = FALSE}
MaxPalindrome <- function() {
  # your code
}
print(MaxPalindrome()) # 906609
```
